---
title: "Data processing for Shahan_et_al_2022"
author: "Vivek Pal"
date: "2026-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
readRenviron("../../.env")
```

## Setting directories variables

These directories does not change for different studies

```{r}
cache_dir <- tools::R_user_dir("scRNAseqPlants", which = "cache")
data_dir <- tools::R_user_dir("scRNAseqPlants", which = "data")
tmp_dir <- tempdir() # ideally fast speed storage such as SSD

fastq_dir <- file.path(cache_dir, "fastq_files")
starsolo_output_dir <- file.path(cache_dir, "starsolo_output")
rds_output_dir <- file.path(data_dir, "rds_output")

dir.create(fastq_dir, recursive = T, showWarnings = F)
dir.create(starsolo_output_dir, recursive = T, showWarnings = F)
dir.create(rds_output_dir, recursive = T, showWarnings = F)

dir.exists(c(tmp_dir, fastq_dir, starsolo_output_dir, rds_output_dir))

pkg_root <- rprojroot::find_package_root_file()
```

## Setting system commands and arguments

Commands can be modified as per system capabilities. We used GNU Parallel for parallel processing of multiple samples. Current values were used on a workstation with 64 threads

```{r}
parallel_arguments <- "-j 10 --colsep ' '"
parallel_arguments_dryrun <- "-j 10 --dry-run --colsep ' '"
prefetch_command <- "prefetch -p --max-size 60G"
fasterq_command <- "./run_fasterq.sh {1} {2} {3}"
star_solo_command <- "./run_starsolo.sh {1} {2} {3} {4} {5}"
```

## Setting study specific variables

```{r}
study_name <- "Shahan_et_al_2022"
# manifest <- read.csv(system.file("extdata", "manifest.csv", package = "scRNAseqPlants"))
manifest <- read.csv(file.path(pkg_root, "inst", "extdata", "manifest.csv"))
meta_data <- manifest[manifest$Study == study_name, ]
samples <- unique(meta_data$GEO_sample)
runs_compacted <- meta_data$SRA_runs
runs <- unlist(strsplit(runs_compacted, ",\\s*"))
genome_keyword <- "at"
chemistry <- "v3"
```

## Runing prefetch 

```{r}
system2("parallel", args = c(parallel_arguments_dryrun, prefetch_command), input = runs)
```

## Running fasterq-dump

```{r}
downloaded <- runs[sapply(runs, function(run) {
    length(list.files(fastq_dir, pattern = paste0("^", run, ".*\\.fastq\\.gz"))) > 0
})]

sapply(runs, function(run) {
    length(list.files(fastq_dir, pattern = paste0("^", run, ".*\\.fastq\\.gz")))
})

if (length(downloaded) > 0) {
    message("runs already downloaded : ", length(downloaded), "/", length(runs), " ", paste(downloaded, collapse = ", "))
}

pending <- runs[!runs %in% downloaded]

if (length(pending) > 0) {
    message("downloading : ", paste(pending, collapse = ", "))
    fasterq_input <- sapply(pending, function(run) {
        paste(run, tmp_dir, fastq_dir)
    })
    system2("parallel", args = c(parallel_arguments_dryrun, fasterq_command), input = fasterq_input)
}

message("files downloaded : ", sapply(runs, function(run) list.files(fastq_dir, pattern = paste0("^", run, ".*\\.fastq\\.gz"))))
```

##  Running star solo 

add description about runs with read1 = 26, those reads were skipped, checked in output also

```{r}
check_read_length <- function(file) {
    cmd <- paste0("gunzip -c ", shQuote(file), " | awk 'NR==2{print length($0); exit}'")
    l <- system(cmd, intern = TRUE)
    return(as.integer(trimws(l)))
}

samples <- samples[!samples %in% c("GSM4625994", "GSM4625995")] # contains rice root protoplasts as spike-ins
solo_script_input <- sapply(samples, FUN = function(sample) {
    sample_runs <- meta_data[meta_data$GEO_sample == sample, "SRA_runs"]
    sample_runs <- unlist(strsplit(sample_runs, ",\\s*"))
    sample_runs_files <- list.files(fastq_dir, pattern = paste0("^(", paste0(sample_runs, collapse = "|"), ").*\\.fastq\\.gz$"), full.names = TRUE)
    sample_runs_files_read_length <- sapply(sample_runs_files, check_read_length)
    reads1 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length == 28]), collapse = ",")
    reads2 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length >= 90]), collapse = ",")
    starsolo_output_prefix <- file.path(starsolo_output_dir, study_name, sample, sample)
    solo_input <- paste0(c(reads1, reads2, genome_keyword, chemistry, starsolo_output_prefix), collapse = " ")
    return(solo_input)
})
system2("parallel", args = c(parallel_arguments_dryrun, star_solo_command), input = solo_script_input)

samples <- c("GSM4625994", "GSM4625995") # contains rice root protoplasts as spike-ins
genome_keyword <- "at+os"
solo_script_input <- sapply(samples, FUN = function(sample) {
    sample_runs <- meta_data[meta_data$GEO_sample == sample, "SRA_runs"]
    sample_runs <- unlist(strsplit(sample_runs, ",\\s*"))
    sample_runs_files <- list.files(fastq_dir, pattern = paste0("^(", paste0(sample_runs, collapse = "|"), ").*\\.fastq\\.gz$"), full.names = TRUE)
    sample_runs_files_read_length <- sapply(sample_runs_files, check_read_length)
    reads1 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length == 28]), collapse = ",")
    reads2 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length >= 90]), collapse = ",")
    starsolo_output_prefix <- file.path(starsolo_output_dir, study_name, sample, sample)
    solo_input <- paste0(c(reads1, reads2, genome_keyword, chemistry, starsolo_output_prefix), collapse = " ")
    return(solo_input)
})
system2("parallel", args = c(parallel_arguments_dryrun, star_solo_command), input = solo_script_input)
```

## make SingleCellExperiment object and metadata for package

```{r}
suppressPackageStartupMessages(library("Matrix"))
suppressPackageStartupMessages(library("SingleCellExperiment"))

filtered_sce_list <- list()
unfiltered_sce_list <- list()
for (sample in samples) {
    sample_metadata <- meta_data[meta_data$GEO_sample == sample, ]

    filtered_matrices <- list(
        "counts"    = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "matrix.mtx")) |> as(Class = "CsparseMatrix"),
        "spliced"   = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "spliced.mtx")) |> as(Class = "CsparseMatrix"),
        "unspliced" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "unspliced.mtx")) |> as(Class = "CsparseMatrix"),
        "ambiguous" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "ambiguous.mtx")) |> as(Class = "CsparseMatrix")
    )

    barcodes <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "barcodes.tsv"))
    barcodes <- cbind(rep(sample, length(barcodes$V1)), barcodes)
    names(barcodes) <- c("Sample", "Barcode")
    features <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "features.tsv"), sep = "\t", quote = "")
    names(features) <- c("ID", "Symbol", "Type")

    filtered_sce <- SingleCellExperiment(filtered_matrices, colData = barcodes, rowData = features, metadata = as.list(sample_metadata))
    print(dim(filtered_sce))

    #############################

    unfiltered_matrices <- list(
        "counts"    = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "matrix.mtx")) |> as(Class = "CsparseMatrix"),
        "spliced"   = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "spliced.mtx")) |> as(Class = "CsparseMatrix"),
        "unspliced" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "unspliced.mtx")) |> as(Class = "CsparseMatrix"),
        "ambiguous" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "ambiguous.mtx")) |> as(Class = "CsparseMatrix")
    )

    raw_barcodes <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "barcodes.tsv"))
    raw_barcodes <- cbind(rep(sample, length(raw_barcodes$V1)), raw_barcodes)
    names(raw_barcodes) <- c("Sample", "Barcode")
    raw_features <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "features.tsv"), sep = "\t", quote = "")
    names(raw_features) <- c("ID", "Symbol", "Type")

    unfiltered_sce <- SingleCellExperiment(unfiltered_matrices, colData = raw_barcodes, rowData = raw_features, metadata = as.list(sample_metadata))
    print(dim(unfiltered_sce))

    filtered_sce_list[[sample]] <- filtered_sce
    unfiltered_sce_list[[sample]] <- unfiltered_sce
}

dir.create(file.path(rds_output_dir, study_name), recursive = TRUE, showWarnings = FALSE)

saveRDS(filtered_sce_list, file = file.path(rds_output_dir, study_name, paste0(study_name, "_filtered_sce_list.rds")), compress = "xz")
saveRDS(unfiltered_sce_list, file = file.path(rds_output_dir, study_name, paste0(study_name, "_unfiltered_sce_list.rds")), compress = "xz")
```

## make metadata.csv for package

```{r}
suppressPackageStartupMessages(library("HubPub"))
suppressPackageStartupMessages(library("ExperimentHubData"))

metadata <- hub_metadata(
    Title = "Shahan_2022_filtered",
    Description = "filtered counts for scRNA-seq data from 19 arabidopsis root tip samples.",
    BiocVersion = package_version("3.24"),
    Genome = "TAIR12",
    SourceType = "FASTQ",
    SourceUrl = "https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE152766",
    SourceVersion = "1.0.0",
    Species = "Arabidopsis thaliana",
    TaxonomyId = as.integer(3702),
    Coordinate_1_based = NA,
    DataProvider = "NCBI",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Shahan_et_al_2022_filtered_sce_list.rds",
    Tags = "scRNA-seq:arabidopsis:root"
)

add_resource(pkg_root, metadata)

metadata <- hub_metadata(
    Title = "Shahan_2022_unfiltered",
    Description = "unfiltered counts for scRNA-seq data from 19 arabidopsis root tip samples.",
    BiocVersion = package_version("3.24"),
    Genome = "TAIR12",
    SourceType = "FASTQ",
    SourceUrl = "https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE152766",
    SourceVersion = "1.0.0",
    Species = "Arabidopsis thaliana",
    TaxonomyId = as.integer(3702),
    Coordinate_1_based = NA,
    DataProvider = "NCBI",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Shahan_et_al_2022_unfiltered_sce_list.rds",
    Tags = "scRNA-seq:arabidopsis:root"
)

add_resource(pkg_root, metadata)
```

## session info
```{r}
sessionInfo()
```
