---
title: "Data processing for Wang_et_al_2021"
author: "Vivek Pal"
date: "2026-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
readRenviron("../../.env")
```

## Setting directories variables

These directories does not change for different studies

```{r}
cache_dir <- tools::R_user_dir("scRNAseqPlants", which = "cache")
data_dir <- tools::R_user_dir("scRNAseqPlants", which = "data")
tmp_dir <- tempdir() # ideally fast speed storage such as SSD

fastq_dir <- file.path(cache_dir, "fastq_files")
starsolo_output_dir <- file.path(cache_dir, "starsolo_output")
rds_output_dir <- file.path(data_dir, "rds_output")

dir.create(fastq_dir, recursive = T, showWarnings = F)
dir.create(starsolo_output_dir, recursive = T, showWarnings = F)
dir.create(rds_output_dir, recursive = T, showWarnings = F)

dir.exists(c(tmp_dir, fastq_dir, starsolo_output_dir, rds_output_dir))

pkg_root <- rprojroot::find_package_root_file()
```

## Setting system commands and arguments

Commands can be modified as per system capabilities. We used GNU Parallel for parallel processing of multiple samples. Current values were used on a workstation with 64 threads

```{r}
parallel_arguments <- "-j 10 --colsep ' '"
parallel_arguments_dryrun <- "-j 10 --dry-run --colsep ' '"
wget_cmd <- paste0("wget -P ", fastq_dir, "/Wang_et_al_2021_tar/ ", "{}")
star_solo_command <- "./run_starsolo.sh {1} {2} {3} {4} {5}"
```

## Setting study specific variables

```{r}
study_name <- "Wang_et_al_2021"
manifest <- read.csv(system.file("extdata", "manifest.csv", package = "scRNAseqPlants"))
meta_data <- manifest[manifest$Study == study_name, ]
samples <- unique(meta_data$SRA_experiment)
runs_compacted <- meta_data$SRA_runs
runs <- unlist(strsplit(runs_compacted, ",\\s*"))
genome_keyword <- "os"
chemistry <- "v3"
```

## downloading originally submitted tar files


```{r}
runs <- meta_data$SRA.run
runs_url <- meta_data$Download_URL
downloaded <- runs[sapply(runs, function(run) {
    file.exists(fastq_dir, pattern = run)
})]
pending <- runs[!runs %in% already_downloaded]
urls <- runs_url[grepl(pattern = paste0(pending, collapse = "|"), runs_url)]

system2("parallel", args = c(parallel_arguments, wget_command), input = urls)
```

## decompress tar to get fastq

```{r}
```


##  Running star solo 

```{r}
samples <- unique(meta_data$Original_sample_id)

check_read_length <- function(file) {
    cmd <- paste0("gunzip -c ", shQuote(file), " | awk 'NR==2{print length($0); exit}'")
    l <- system(cmd, intern = TRUE)
    return(as.integer(trimws(l)))
}

solo_script_input <- sapply(samples, FUN = function(sample) {
    exp <- meta_data[meta_data$Original_sample_id == sample, "SRA_experiment"]
    sample_runs_files <- list.files(fastq_dir, pattern = paste0("^(", paste0(sample, collapse = "|"), ").*\\.fastq\\.gz$"), full.names = TRUE)
    sample_runs_files_read_length <- sapply(sample_runs_files, check_read_length)
    # print(sample_runs_files_read_length) # lengths are consistent
    reads1 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length == 28]), collapse = ",")
    reads2 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length > 90]), collapse = ",")
    starsolo_output_prefix <- file.path(starsolo_output_dir, study, exp, exp)
    solo_input <- paste0(c(reads1, reads2, genome_keyword, chemistry, starsolo_output_prefix), collapse = " ")
    return(solo_input)
})

system2("parallel", args = c(parallel_arguments, star_solo_command), input = solo_script_input)
```

## make SingleCellExperiment object 

```{r}
```

## make metadata.csv for package

```{r}
suppressPackageStartupMessages(library("HubPub"))
suppressPackageStartupMessages(library("ExperimentHubData"))

metadata <- hub_metadata(
    Title = "Wang_2021_filtered",
    Description = "filtered counts for scRNA-seq data from 22 rice samples for cultivar Nip, tissues contain control and treated leaf, root and seedling.",
    BiocVersion = package_version("3.24"),
    Genome = "AGIS1.0",
    SourceType = "FASTQ",
    SourceUrl = "https://download.cncb.ac.cn/gsa/CRA004082/",
    SourceVersion = "1.0.0",
    Species = "Oryza sativa",
    TaxonomyId = as.integer(4530),
    Coordinate_1_based = NA,
    DataProvider = "NGDC",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Wang_et_al_2021_filtered_sce_list.rds",
    Tags = "scRNA-seq:arabidopsis:root:leaf:seedling"
)

add_resource(pkg_root, metadata)

metadata <- hub_metadata(
    Title = "Wang_2021_unfiltered",
    Description = "unfiltered counts for scRNA-seq data from 22 rice samples for cultivar Nip, tissues contain control and treated leaf, root and seedling.",
    BiocVersion = package_version("3.24"),
    Genome = "AGIS1.0",
    SourceType = "FASTQ",
    SourceUrl = "https://download.cncb.ac.cn/gsa/CRA004082/",
    SourceVersion = "1.0.0",
    Species = "Oryza sativa",
    TaxonomyId = as.integer(4530),
    Coordinate_1_based = NA,
    DataProvider = "NGDC",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Wang_et_al_2021_unfiltered_sce_list.rds",
    Tags = "scRNA-seq:arabidopsis:root:leaf:seedling"
)

add_resource(pkg_root, metadata)
```

## session info
```{r}
sessionInfo()
```
