---
title: "Data processing for Wang_et_al_2021"
author: "Vivek Pal"
date: "2026-05-30"
output: html_document
---

```{r setup, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup system variables for external tools

We used a .env file to setup external tools such as STAR, CellRanger(for whitelist file) and genome indexes location. An .env.example file is provided for users who want to reproduce the steps in this workflow.

```{r}
env_file <- file.path(rprojroot::find_package_root_file(), ".env")
if (file.exists(env_file)) {
    readRenviron(env_file)
} else {
    message("`.env` file not found at: ", env_file, "\nSystem environment variables will not be used.")
}
```

## Setting directories as variables

These directories does not change for different studies

```{r}
cache_dir <- tools::R_user_dir("scRNAseqPlants", which = "cache")
data_dir <- tools::R_user_dir("scRNAseqPlants", which = "data")
tmp_dir <- tempdir() # ideally fast speed storage such as SSD

fastq_dir <- file.path(cache_dir, "fastq_files")
starsolo_output_dir <- file.path(cache_dir, "starsolo_output")
rds_output_dir <- file.path(data_dir, "rds_output")

dir.create(fastq_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(starsolo_output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(rds_output_dir, recursive = TRUE, showWarnings = FALSE)

dir.exists(c(tmp_dir, fastq_dir, starsolo_output_dir, rds_output_dir))

pkg_root <- rprojroot::find_package_root_file()
```

## Setting system commands and arguments

Commands can be modified as per system capabilities. We used GNU Parallel for parallel processing of multiple samples. Current values were used on a workstation with 64 threads. In this document we used parallel_arguments_dryrun in system2() calls, for actual execution use parallel_arguments. run_starsolo.sh is included in the inst/scripts directory.

```{r}
parallel_arguments <- "-j 10 --colsep '\\|'"
parallel_arguments_dryrun <- "-j 10 --dry-run --colsep '\\|'"
wget_cmd <- paste0("wget -P ", fastq_dir, "/Wang_et_al_2021_tar/ ", "{}")
untar_cmd <- paste0("tar -xvf ", "{}", " -C ", fastq_dir)
star_solo_command <- "./run_starsolo.sh {1} {2} {3} {4} {5}"
```

## Setting study specific variables

```{r, eval=FALSE}
study_name <- "Wang_et_al_2021"
manifest <- read.csv(system.file("extdata", "manifest.csv", package = "scRNAseqPlants"))
meta_data <- manifest[manifest$Study == study_name, ]
samples <- unique(meta_data$SRA_experiment)
samples_name <- unique(meta_data$Original_sample_id)
runs_compacted <- meta_data$SRA_runs
runs <- unlist(strsplit(runs_compacted, ",\\s*"))
genome_keyword <- "os"
chemistry <- "v3"
```

## Downloading originally submitted tar files

We downloaded the tarballs using ftp links available at https://ngdc.cncb.ac.cn/gsa/browse/CRA004082
```{r, eval=FALSE}
runs_url_compacted <- meta_data$Download_URL
runs_url <- unlist(strsplit(runs_url_compacted, ",\\s*"))
downloaded <- runs[sapply(runs, function(run) {
    length(list.files(file.path(fastq_dir, "Wang_et_al_2021_tar"), pattern = run)) > 0
})]
pending <- runs[!runs %in% downloaded]
urls <- runs_url[grepl(pattern = paste0(pending, collapse = "|"), runs_url)]

system2("parallel", args = c(parallel_arguments_dryrun, wget_cmd), input = urls)
```

## Decompress tar to get fastq files

```{r, eval=FALSE}
tars <- list.files(path = file.path(fastq_dir, "Wang_et_al_2021_tar"), full.names = TRUE)

system2("parallel", args = c(parallel_arguments_dryrun, untar_cmd), input = tars)
```


##  Running star solo 

```{r, eval=FALSE}
check_read_length <- function(file) {
    cmd <- paste0("gunzip -c ", shQuote(file), " | awk 'NR==2{print length($0); exit}'")
    l <- system(cmd, intern = TRUE)
    return(as.integer(trimws(l)))
}

solo_script_input <- sapply(samples_name, FUN = function(sample) {
    exp <- meta_data[meta_data$Original_sample_id == sample, "SRA_experiment"]
    sample_runs_files <- list.files(fastq_dir, pattern = paste0("^(", paste0(sample, collapse = "|"), ").*\\.fastq\\.gz$"), full.names = TRUE)
    sample_runs_files_read_length <- sapply(sample_runs_files, check_read_length)
    # print(sample_runs_files_read_length) # consistent lengths
    reads1 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length == 28]), collapse = ",")
    reads2 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length >= 90]), collapse = ",")
    starsolo_output_prefix <- file.path(starsolo_output_dir, study_name, exp, exp)
    solo_input <- paste0(c(reads1, reads2, genome_keyword, chemistry, starsolo_output_prefix), collapse = "|")
    return(solo_input)
})

system2("parallel", args = c(parallel_arguments_dryrun, star_solo_command), input = solo_script_input)
```

## Make SingleCellExperiment object and save RDS for package

```{r, eval=FALSE}
suppressPackageStartupMessages(library("Matrix"))
suppressPackageStartupMessages(library("SingleCellExperiment"))

filtered_sce_list <- list()
unfiltered_sce_list <- list()
for (sample in samples) {
    sample_metadata <- meta_data[meta_data$SRA_experiment == sample, ]

    filtered_matrices <- list(
        "counts"    = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "matrix.mtx")) |> as(Class = "CsparseMatrix"),
        "spliced"   = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "spliced.mtx")) |> as(Class = "CsparseMatrix"),
        "unspliced" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "unspliced.mtx")) |> as(Class = "CsparseMatrix"),
        "ambiguous" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "ambiguous.mtx")) |> as(Class = "CsparseMatrix")
    )

    barcodes <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "barcodes.tsv"))
    barcodes <- cbind(rep(sample, length(barcodes$V1)), barcodes)
    names(barcodes) <- c("Sample", "Barcode")
    features <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "features.tsv"), sep = "\t", quote = "")
    names(features) <- c("ID", "Symbol", "Type")

    filtered_sce <- SingleCellExperiment(filtered_matrices, colData = barcodes, rowData = features, metadata = as.list(sample_metadata))
    print(dim(filtered_sce))

    #############################

    unfiltered_matrices <- list(
        "counts"    = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "matrix.mtx")) |> as(Class = "CsparseMatrix"),
        "spliced"   = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "spliced.mtx")) |> as(Class = "CsparseMatrix"),
        "unspliced" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "unspliced.mtx")) |> as(Class = "CsparseMatrix"),
        "ambiguous" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "ambiguous.mtx")) |> as(Class = "CsparseMatrix")
    )

    raw_barcodes <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "barcodes.tsv"))
    raw_barcodes <- cbind(rep(sample, length(raw_barcodes$V1)), raw_barcodes)
    names(raw_barcodes) <- c("Sample", "Barcode")
    raw_features <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "features.tsv"), sep = "\t", quote = "")
    names(raw_features) <- c("ID", "Symbol", "Type")

    unfiltered_sce <- SingleCellExperiment(unfiltered_matrices, colData = raw_barcodes, rowData = raw_features, metadata = as.list(sample_metadata))
    print(dim(unfiltered_sce))

    filtered_sce_list[[sample]] <- filtered_sce
    unfiltered_sce_list[[sample]] <- unfiltered_sce
}

dir.create(file.path(rds_output_dir, study_name), recursive = TRUE, showWarnings = FALSE)

saveRDS(filtered_sce_list, file = file.path(rds_output_dir, study_name, paste0(study_name, "_filtered_sce_list.rds")), compress = "xz")
saveRDS(unfiltered_sce_list, file = file.path(rds_output_dir, study_name, paste0(study_name, "_unfiltered_sce_list.rds")), compress = "xz")
```

## Make metadata.csv for package

```{r, eval=FALSE}
suppressPackageStartupMessages(library("HubPub"))

metadata <- hub_metadata(
    Title = "Wang_2021_filtered",
    Description = "filtered counts for scRNA-seq data from 22 rice samples for cultivar Nip, tissues contain control and treated leaf, root and seedling.",
    BiocVersion = BiocManager::version(),
    Genome = "AGIS1.0",
    SourceType = "FASTQ",
    SourceUrl = "https://download.cncb.ac.cn/gsa/CRA004082/",
    SourceVersion = "1.0.0",
    Species = "Oryza sativa",
    TaxonomyId = as.integer(4530),
    Coordinate_1_based = NA,
    DataProvider = "NGDC",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Wang_et_al_2021_filtered_sce_list.rds",
    Tags = "scRNA-seq:rice:root:leaf:seedling"
)

add_resource(pkg_root, metadata)

metadata <- hub_metadata(
    Title = "Wang_2021_unfiltered",
    Description = "unfiltered counts for scRNA-seq data from 22 rice samples for cultivar Nip, tissues contain control and treated leaf, root and seedling.",
    BiocVersion = BiocManager::version(),
    Genome = "AGIS1.0",
    SourceType = "FASTQ",
    SourceUrl = "https://download.cncb.ac.cn/gsa/CRA004082/",
    SourceVersion = "1.0.0",
    Species = "Oryza sativa",
    TaxonomyId = as.integer(4530),
    Coordinate_1_based = NA,
    DataProvider = "NGDC",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Wang_et_al_2021_unfiltered_sce_list.rds",
    Tags = "scRNA-seq:rice:root:leaf:seedling"
)

add_resource(pkg_root, metadata)
```

## session info
Note: packages below are loaded conditionally in eval=FALSE chunks.
sessionInfo() reflects the minimal environment for knitting this document.
```{r}
sessionInfo()
```
