---
title: "Data processing for Zhang_et_al_2019"
author: "Vivek Pal"
date: "2026-05-30"
output: html_document
---

```{r setup, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup system variables for external tools

We used a .env file to setup external tools such as SRA toolkit, pigz(for fast compression), STAR, CellRanger(for whitelist file) and genome indexes location. An .env.example file is provided for users who want to reproduce the steps in this workflow.

```{r}
env_file <- file.path(rprojroot::find_package_root_file(), ".env")
if (file.exists(env_file)) {
    readRenviron(env_file)
} else {
    message("`.env` file not found at: ", env_file, "\nSystem environment variables will not be used.")
}
```

## Setting directories as variables

These directories does not change for different studies

```{r}
cache_dir <- tools::R_user_dir("scRNAseqPlants", which = "cache")
data_dir <- tools::R_user_dir("scRNAseqPlants", which = "data")
tmp_dir <- tempdir() # ideally fast speed storage such as SSD

fastq_dir <- file.path(cache_dir, "fastq_files")
starsolo_output_dir <- file.path(cache_dir, "starsolo_output")
rds_output_dir <- file.path(data_dir, "rds_output")

dir.create(fastq_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(starsolo_output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(rds_output_dir, recursive = TRUE, showWarnings = FALSE)

dir.exists(c(tmp_dir, fastq_dir, starsolo_output_dir, rds_output_dir))

pkg_root <- rprojroot::find_package_root_file()
```

## Setting system commands and arguments

Commands can be modified as per system capabilities. We used GNU Parallel for parallel processing of multiple samples. Current values were used on a workstation with 64 threads. In this document we used parallel_arguments_dryrun in system2() calls, for actual execution use parallel_arguments. run_fasterq.sh and run_starsolo.sh are included in the inst/scripts directory

```{r}
parallel_arguments <- "-j 10 --colsep ' '"
parallel_arguments_dryrun <- "-j 10 --dry-run --colsep ' '"
prefetch_command <- paste0(file.path(Sys.getenv("SRATOOLKIT_BIN"), "prefetch"), " -p --max-size 60G")
fasterq_command <- "./run_fasterq.sh {1} {2} {3}"
star_solo_command <- "./run_starsolo.sh {1} {2} {3} {4} {5}"
```

## Setting study specific variables

```{r}
study_name <- "Zhang_et_al_2019"
manifest <- read.csv(system.file("extdata", "manifest.csv", package = "scRNAseqPlants"))
meta_data <- manifest[manifest$Study == study_name, ]
samples <- unique(meta_data$SRA_experiment)
runs_compacted <- meta_data$SRA_runs
runs <- unlist(strsplit(runs_compacted, ",\\s*"))
genome_keyword <- "at"
chemistry <- "v2"
```

## Running prefetch 

We downloaded the .sra run files using NCBI's SRA Toolkit. Path to sra toolkit should be specified in .env file

```{r, eval=FALSE}
system2("parallel", args = c(parallel_arguments_dryrun, prefetch_command), input = runs)
```

## Running fasterq-dump

When setup correctly, fasterq-dump should automatically find the prefetched runs.

```{r, eval=FALSE}
downloaded <- runs[sapply(runs, function(run) {
    length(list.files(fastq_dir, pattern = paste0("^", run, ".*\\.fastq\\.gz"))) > 0
})]

sapply(runs, function(run) {
    length(list.files(fastq_dir, pattern = paste0("^", run, ".*\\.fastq\\.gz")))
})

if (length(downloaded) > 0) {
    message("runs already downloaded : ", length(downloaded), "/", length(runs), " ", paste(downloaded, collapse = ", "))
}

pending <- runs[!runs %in% downloaded]

if (length(pending) > 0) {
    message("downloading : ", paste(pending, collapse = ", "))
    fasterq_input <- sapply(pending, function(run) {
        paste(run, tmp_dir, fastq_dir)
    })
    system2("parallel", args = c(parallel_arguments_dryrun, fasterq_command), input = fasterq_input)
}

message("files downloaded : ", sapply(runs, function(run) list.files(fastq_dir, pattern = paste0("^", run, ".*\\.fastq\\.gz"))))
```

##  Running star solo 

```{r, eval=FALSE}
check_read_length <- function(file) {
    cmd <- paste0("gunzip -c ", shQuote(file), " | awk 'NR==2{print length($0); exit}'")
    l <- system(cmd, intern = TRUE)
    return(as.integer(trimws(l)))
}

solo_script_input <- sapply(samples, FUN = function(sample) {
    sample_runs <- meta_data[meta_data$SRA_experiment == sample, "SRA_runs"]
    sample_runs <- unlist(strsplit(sample_runs, ",\\s*"))
    sample_runs_files <- list.files(fastq_dir, pattern = paste0("^(", paste0(sample_runs, collapse = "|"), ").*\\.fastq\\.gz$"), full.names = TRUE)
    sample_runs_files_read_length <- sapply(sample_runs_files, check_read_length)
    # print(sample_runs_files_read_length) # consistent lengths
    reads1 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length == 26]), collapse = ",")
    reads2 <- paste0(names(sample_runs_files_read_length[sample_runs_files_read_length >= 90]), collapse = ",")
    starsolo_output_prefix <- file.path(starsolo_output_dir, study_name, sample, sample)
    solo_input <- paste0(c(reads1, reads2, genome_keyword, chemistry, starsolo_output_prefix), collapse = " ")
    return(solo_input)
})

system2("parallel", args = c(parallel_arguments_dryrun, star_solo_command), input = solo_script_input)
```

## Make SingleCellExperiment object and save RDS for package

```{r, eval=FALSE}
suppressPackageStartupMessages(library("Matrix"))
suppressPackageStartupMessages(library("SingleCellExperiment"))

filtered_sce_list <- list()
unfiltered_sce_list <- list()
for (sample in samples) {
    sample_metadata <- meta_data[meta_data$SRA_experiment == sample, ]

    filtered_matrices <- list(
        "counts"    = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "matrix.mtx")) |> as(Class = "CsparseMatrix"),
        "spliced"   = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "spliced.mtx")) |> as(Class = "CsparseMatrix"),
        "unspliced" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "unspliced.mtx")) |> as(Class = "CsparseMatrix"),
        "ambiguous" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "filtered", "ambiguous.mtx")) |> as(Class = "CsparseMatrix")
    )

    barcodes <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "barcodes.tsv"))
    barcodes <- cbind(rep(sample, length(barcodes$V1)), barcodes)
    names(barcodes) <- c("Sample", "Barcode")
    features <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "filtered", "features.tsv"), sep = "\t", quote = "")
    names(features) <- c("ID", "Symbol", "Type")

    filtered_sce <- SingleCellExperiment(filtered_matrices, colData = barcodes, rowData = features, metadata = as.list(sample_metadata))
    print(dim(filtered_sce))

    #############################

    unfiltered_matrices <- list(
        "counts"    = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "matrix.mtx")) |> as(Class = "CsparseMatrix"),
        "spliced"   = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "spliced.mtx")) |> as(Class = "CsparseMatrix"),
        "unspliced" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "unspliced.mtx")) |> as(Class = "CsparseMatrix"),
        "ambiguous" = readMM(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Velocyto", "raw", "ambiguous.mtx")) |> as(Class = "CsparseMatrix")
    )

    raw_barcodes <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "barcodes.tsv"))
    raw_barcodes <- cbind(rep(sample, length(raw_barcodes$V1)), raw_barcodes)
    names(raw_barcodes) <- c("Sample", "Barcode")
    raw_features <- read.table(file.path(starsolo_output_dir, study_name, sample, paste0(sample, "Solo.out"), "Gene", "raw", "features.tsv"), sep = "\t", quote = "")
    names(raw_features) <- c("ID", "Symbol", "Type")

    unfiltered_sce <- SingleCellExperiment(unfiltered_matrices, colData = raw_barcodes, rowData = raw_features, metadata = as.list(sample_metadata))
    print(dim(unfiltered_sce))

    filtered_sce_list[[sample]] <- filtered_sce
    unfiltered_sce_list[[sample]] <- unfiltered_sce
}

dir.create(file.path(rds_output_dir, study_name), recursive = TRUE, showWarnings = FALSE)

saveRDS(filtered_sce_list, file = file.path(rds_output_dir, study_name, paste0(study_name, "_filtered_sce_list.rds")), compress = "xz")
saveRDS(unfiltered_sce_list, file = file.path(rds_output_dir, study_name, paste0(study_name, "_unfiltered_sce_list.rds")), compress = "xz")
```

## Make metadata.csv for package

```{r, eval=FALSE}
suppressPackageStartupMessages(library("HubPub"))

metadata <- hub_metadata(
    Title = "Zhang_2019_filtered",
    Description = "filtered counts for scRNA-seq data from 1 arabidopsis root tip sample.",
    BiocVersion = BiocManager::version(),
    Genome = "TAIR12",
    SourceType = "FASTQ",
    SourceUrl = "https://trace.ncbi.nlm.nih.gov/Traces/index.html?view=run_browser&acc=SRR8485805",
    SourceVersion = "1.0.0",
    Species = "Arabidopsis thaliana",
    TaxonomyId = as.integer(3702),
    Coordinate_1_based = NA,
    DataProvider = "NCBI",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Zhang_et_al_2019_filtered_sce_list.rds",
    Tags = "scRNA-seq:arabidopsis:root"
)

add_resource(pkg_root, metadata)

metadata <- hub_metadata(
    Title = "Zhang_2019_unfiltered",
    Description = "unfiltered counts for scRNA-seq data from 1 arabidopsis root tip sample.",
    BiocVersion = BiocManager::version(),
    Genome = "TAIR12",
    SourceType = "FASTQ",
    SourceUrl = "https://trace.ncbi.nlm.nih.gov/Traces/index.html?view=run_browser&acc=SRR8485805",
    SourceVersion = "1.0.0",
    Species = "Arabidopsis thaliana",
    TaxonomyId = as.integer(3702),
    Coordinate_1_based = NA,
    DataProvider = "NCBI",
    Maintainer = "Vivek Pal <vivek86_sit@jnu.ac.in>",
    RDataClass = "list",
    DispatchClass = "Rds",
    Location_Prefix = "https://zenodo.org/",
    RDataPath = "records/20976557/files/Zhang_et_al_2019_unfiltered_sce_list.rds",
    Tags = "scRNA-seq:arabidopsis:root"
)

add_resource(pkg_root, metadata)
```

## session info
Note: packages below are loaded conditionally in eval=FALSE chunks.
sessionInfo() reflects the minimal environment for knitting this document.
```{r}
sessionInfo()
```
